Follow-up Samples

### 2017-02-07 submission/manuscript_outline_20161024.Rmd
explain CRC associated OTUs better -Results

***Bacterial Community and FIT***
Based on the thetayc distance metric, comparing the initial to the follow up samples, there was a significant difference between the adenoma and carcinoma groups (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. There was also a significant difference in change in FIT between initial and follow up samples (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The whole community structure before and after surgery was visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When all initial and follow up samples were compared to each other there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, or evenness after correction for multiple comparisons [Table S1].  There was also no significant difference between initial and follow up samples for any single OTU [Figure S1].

***Previously Associated Cancer Bacteria***
We next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  First, the carcinoma samples showed a significant difference between initial and follow up samples for *Peptostreptococcus stomatis* (P-value = `r format(main_crc_bugs_summary["pept", "crc_BH"], digits = 3)`), *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`) whereas there were no significant differences in any of these OTUs in the adenoma samples. Second, when these OTUs were present, there was a clear magnitude difference based on whether they were from adenoma or carcinoma [Figure 2].  However, only a small percentage of those with adenoma or carcinoma were positive for any of these OTUs.

>>>>

***Bacterial Community and FIT***
We first wanted to test whether there were any broad differences between initial and follow up samples based on lesion being either adenoma or carcinoma.  What we found was that the bacterial community in those with carcinoma were more dissimilar (as measured by thetayc) to their initial sample then those with adenoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. We also found that there were larger changes fecal blood (measured by FIT) for those with carcinoma versus adenoma (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  The broad shift in bacterial community structure before and after surgery was visualized on NMDS graphs for both adenoma [Figure1c] (PERMANOVA = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) and carcinoma [Figure 1d] (PERMANOVA = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  Interestingly, when initial and follow up samples were compared to each other, regardless of whether they were adenoma or carcinoma (lesion), there was no significant overall difference between them (PERMANOVA = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  When investigating more broad alpha diversity metrics there was no difference found between initial and follow up samples for lesion, adenoma, or carcinoma for any metric [Table S1].  We also observed that there was no difference in OTU relative abundance between initial and follow up samples for lession, adenoma only, or carcinoma only [Figure S1].

***Previously Associated Cancer Bacteria***
Previous literature has suggested that a number of oral microbes may be important in CRC pathogenesis [@flynn_metabolic_2016].  So we next examined whether there were differences in previously well described carcinoma associated OTUs.  These included the OTUs that aligned with *Porphyromonas asaccharolytica* (`r unique(filter(crc_bug_info, tax_id == "Porphyromonas")[, "otu"])`), *Fusobacterium nucleatum* (`r unique(filter(crc_bug_info, tax_id == "Fusobacterium")[, "otu"])`), *Parvimonas micra* (`r unique(filter(crc_bug_info, tax_id == "Parvimonas")[, "otu"])`), and *Peptostreptococcus stomatis* (`r unique(filter(crc_bug_info, tax_id == "Peptostreptococcus")[, "otu"])`).  There was a difference in initial and follow up samples for lesion and carcinoma for relative abundance of *Parvimonas micra* (P-value = `r format(main_crc_bugs_summary["parv", "crc_BH"], digits = 3)`), and *Porphyromonas asaccharolytica* (P-value = `r format(main_crc_bugs_summary["porp", "crc_BH"], digits = 3)`). In contrast, there was no differences relative abundance in any of these OTUs for those with adenoma [Figure 2]. We also observed that only a small percentage of those with adenoma or carcinoma were positive or had an appreciable relative abundance of these OTUs [Figure 2].
